---
title: "EEP analysis good copy"
output: html_document
date: "2026-03-03"
---


load libraries 
```{r}
library(tidyverse) 
library(readxl) 
library(lme4) #for fancy modelling 
library(lmerTest) #testing lmer
library(scales) #not using yet
library(DHARMa) #using this!
library(StatisticalModels) #for r2 glmer however???? it doesnt work :(
library(ggeffects) #not using yet 
library(performance) #not usiung yet 
library(ggthemes) #for pretty grpahs
library(emmeans) # for tukesy test
```

load data 
```{r}
ferndata<-read_excel("ferndata.xlsx")

ferndata %>% group_by(Substrate) %>% summarise(meanfrond = mean(frondlength))
```

data wrangling 

```{r}
ferndata<- ferndata %>% mutate(fertileratio= fertilefrond/numberfrond, widthlengthratio=bladewidth/bladelength) #create new columns 
ferndata$Substrate<- as.factor(ferndata$Substrate)
ferndata<- ferndata %>% mutate(log_frondlength=log(frondlength), log_bladelength=log(bladelength), log_bladewidth=log(bladewidth), log_stipelength=log(stipelength))
```

 ANOVA with random effects for frond length
 
 what does rand do 
```{r}
test3<-lmer(log_frondlength~Substrate+(1|Species), data=ferndata)
anova(test3) 
#p and f value stronger than before 
rand(test3)
#not quite as good fit but still works well
#cant remember what this does 
summary(test3)
#look at the intercept and substrate rocks and soil under fixed effects,
#look at estimate, and from that interpret the percent change from each substrate 

#get r2
R2GLMER(test3)


#DHARMA time 
simulateResiduals(fittedModel = test3, plot=T)
```
Tukeys test 

```{r}
tukeytest<- emmeans(test3, list(pairwise~Substrate), adjust ="tukey")
tukeytest
```

 back transform means (were logged)
```{r}
exp(4.46)
#86.48751 for concrete

exp(5.41)
#223.6316 for rock

exp(5.99)
#399.4146 for soil 
```
 
 
 
figure for frond length vs substrate 

```{r}
#new table
sum_table <- ferndata %>% group_by(Substrate) %>% summarise (mean_frondlength = mean(frondlength), sdev = sd(frondlength), se_length = sd(frondlength)/sqrt(n()))

#figure 
frondsub<- ggplot()+
  geom_point(data =ferndata, aes(x=Substrate , y=frondlength), shape=21)+
  geom_point(data = ferndata, aes(x=Substrate , y=frondlength), stat="summary", fun="mean", size=4)+
  geom_errorbar(data=sum_table, aes(x=Substrate, ymin=(mean_frondlength-se_length), ymax=(mean_frondlength+se_length)), width=0.1)+
  labs(x= "Substrate", y= "Frond Length (mm)")+
  theme_few()+
  theme(legend.position = "none")

frondsub
ggsave("frondsub2.png", width = 10, height = 7, dpi = 600)

```





stats for fertile frond ratio: 

```{r}
#anova for fertile frond ratio 

fertrat<- lm(fertileratio~Substrate, data = ferndata)

par(mfrow = c(1,2))  
hist(fertrat$residuals)   # Makes histogram of residuals  
plot(fertrat, which = 2)   # Makes Q-Q plot
par(mfrow = c(1,1))
#normality is fine 

# Checking homoscedasticity 
plot(fertrat, which = 1)  # Makes residuals VS fitted plot
#kinda violated - run welch f test instead? 


anova(fertrat)


test4<-lmer(fertileratio~Substrate+(1|Species), data=ferndata, REML=TRUE) #make mixed effects test 
anova(test4) # run anova 
simulateResiduals(fittedModel = test4, plot=T) #check assumptions 
R2GLMER(test4)

```

 
figure for frond length

```{r}
#new table
sum_table2 <- ferndata %>% group_by(Substrate) %>% summarise (mean_fertrat = mean(fertileratio), sdev_fertrat = sd(fertileratio), se_fertrat = sd(fertileratio)/sqrt(n()))

#figure 
fertilegraph<- ggplot()+
  geom_violin(data =ferndata, aes(x=Substrate , y=fertileratio))+
  geom_point(data = ferndata, aes(x=Substrate , y=fertileratio), stat="summary", fun="mean", size =3)+
  geom_errorbar(data=sum_table2, aes(x=Substrate, ymin=(mean_fertrat-se_fertrat), ymax=(mean_fertrat+se_fertrat)), width=0.1)+
  scale_x_discrete(labels=c("concrete"="Artificial Rock", "rock"="Natural Rock", "soil"="Soil"))+
  labs(x= "Substrate", y= "Ratio of fertile fronds to total fronds")+
  ylim(0,1)+
  theme_few()+
  theme(legend.position = "none")

fertilegraph

ggsave("fertilegraph.png", width = 9, height = 6, dpi = 600)
```



making a graph shwoing species distribution in each substrate 

```{r}
abundsp<- ferndata %>% group_by(Substrate) %>% count(Species)

abundsp<-abundsp %>% mutate(abundance=n, totalab=sum(abundance), relabund=abundance/totalab)



#CHANGE NAMES OF GROUP LABELS 
RAbar<-ggplot(data=abundsp, aes(y=relabund, x=Species, fill=Substrate))+
  geom_bar(stat="identity")+
  theme_few()+
  facet_wrap(~Substrate, ncol=1)+
  theme(axis.text.x = element_text(angle = 90, size = 10, hjust=1, vjust=0.5), 
          axis.title = element_text(size=15),
          legend.position = "none")+
  labs(y ="Relative Abundance")+
  scale_fill_manual(values=c("Wall"="grey60", "Soil"="#855D38", "Rock"="#853838"))

RAbar

ggsave("RAbar.png", width = 9, height = 9, dpi = 600)
```




now plot for blade width: 
- im thinking a lot of this would be controlled for by species

```{r}
widthmodel<-lmer(bladewidth~Substrate + (1|Species), data=ferndata)

anova(widthmodel)

simulateResiduals(fittedModel = widthmodel, plot=T)

# try logging data - voilates lots of assumptions 


logwidthmodel<-lmer(log_bladewidth ~Substrate +(1|Species), data=ferndata)

anova(logwidthmodel)

R2GLMER(logwidthmodel)

simulateResiduals(fittedModel = logwidthmodel, plot=T)

```

Tukeys test
```{r}
tukeytest3<- emmeans(logwidthmodel, list(pairwise~Substrate), adjust ="tukey")
tukeytest3
```

blade width graph: 


```{r}
#new table
widthtable <- ferndata %>% group_by(Substrate) %>% summarise (mean_bladewidth = mean(bladewidth), sdev_bladewidth= sd(bladewidth), se_bladewidth = sd(bladewidth)/sqrt(n()))

#figure 
frondsub<- ggplot()+
  geom_point(data =ferndata, aes(x=Substrate , y=bladewidth), shape=21)+
  geom_point(data = ferndata, aes(x=Substrate , y=bladewidth), stat="summary", fun="mean", size=4)+
  geom_errorbar(data=widthtable, aes(x=Substrate, ymin=(mean_bladewidth-se_bladewidth), ymax=(mean_bladewidth+se_bladewidth)), width=0.1)+
  labs(x= "Substrate", y= "Blade Width (mm)")+
  theme_few()+
  theme(legend.position = "none")

frondsub
ggsave("bladewidth.png", width = 10, height = 7, dpi = 600)
```



stipe length vs substrate 


```{r}
stipemod<-lmer(log_stipelength ~ Substrate + (1|Species), data=ferndata)
anova(stipemod)

R2GLMER(stipemod)

simulateResiduals(fittedModel = stipemod, plot=T)

```


tukey test for stipe: 
```{r}
tukeyteststipe<- emmeans(stipemod, list(pairwise~Substrate), adjust ="tukey")

tukeyteststipe 
```


