---
title: "EEP analysis good copy"
output: html_document
date: "2026-03-03"
---


load libraries 
```{r}
library(tidyverse) 
library(readxl) 
library(lme4) #for fancy modelling 
library(lmerTest) #testing lmer
library(scales) #not using yet
library(DHARMa) #using this!
library(StatisticalModels) #for r2 glmer however???? it doesnt work :(
library(ggeffects) #not using yet 
library(performance) #not usiung yet 
library(ggthemes) #for pretty grpahs
library(emmeans) # for tukesy test
library(ggpredict)
```

load data 
```{r}
ferndata2<-read_excel("ferndata.xlsx")

ferndata2 %>% group_by(Substrate) %>% summarise(meanfrond = mean(frondlength))


```

is soil 372.449 more than rock? 
```{r}
585.46-383
```



data wrangling 

```{r}
ferndata2<- ferndata %>% mutate(fertileratio= fertilefrond/numberfrond, widthlengthratio=bladewidth/bladelength) #create new columns 
ferndata2$Substrate<- as.factor(ferndata2$Substrate)
ferndata2<- ferndata2 %>% mutate(log_frondlength=log(frondlength), log_bladelength=log(bladelength), log_bladewidth=log(bladewidth), log_stipelength=log(stipelength))
```



 ANOVA with random effects for frond length
 
 what does rand do 
```{r}
test3<-lmer(log_frondlength~Substrate+(1|Species), data=ferndata2)
anova(test3) 
#p and f value stronger than before 
rand(test3)
#not quite as good fit but still works well
#cant remember what this does 
summary(test3)
#look at the intercept and substrate rocks and soil under fixed effects,
#look at estimate, and from that interpret the percent change from each substrate 

#get r2
R2GLMER(test3)

confint(test3)


#DHARMA time 
simulateResiduals(fittedModel = test3, plot=T)


```


FROND LENGTH
```{r}
#maths

#rock mean 
exp(5.3691)
#ROCK CI
exp(4.8388668)
exp(5.9251576)

#soil mean 
exp(5.3691+0.5510)
#soil ci 
exp(5.3691+0.1138408)
exp(5.3691+1.0121938)

#wall mean 
exp(5.3691-0.9291)
#wall ci 
exp(5.3691-1.5068913)
exp(5.3691-0.3663185)


# soil percentage higher than rock 
#new value - original value / origional valuie * 100 
((372.449 - 214.6696)/ 214.6696)*100

```

mean for rock: 214.6696
95% CI for rock: 126.3261, 374.3374 

mean for soil: 372.449
soil ci: 240.5531, 590.6915 

mean for wall: 84.77494
wall ci: 47.5703, 148.8265

mean for soil is 73.5 % higher than rock (Liner Mixed Effect Model t=2.439, p<0.001).
mean for wall is significantly lower than rock (T = -3.206, P = 0.00181 )

table: means for each and confidence intervals 

 
figure for frond length vs substrate 

```{r}
#new table
sum_table <- ferndata2 %>% group_by(Substrate) %>% summarise (mean_frondlength = mean(frondlength), sdev = sd(frondlength), se_length = sd(frondlength)/sqrt(n()))

frondlen_df<- data.frame(Substrate = c("Rock", "Soil", "Wall"), mean_frondlength=c(214.6696, 372.449, 84.77494), lowci=c(126.3261, 240.5531, 47.5703), highci=c(374.3374, 590.6915, 148.8265))

#figure 
frondsub<- ggplot()+
  geom_violin(data =ferndata2, aes(x=Substrate , y=frondlength), blade = 1.5)+
  geom_boxplot(data = ferndata, aes(x=Substrate , y=frondlength),width=0.04, color="#55B03C", alpha=0.2)+
  labs(x= "Substrate", y= "Frond Length (mm)")+
  theme_few()+
  theme(legend.position = "none")

frondsub
ggsave("frondsub2.png", width = 9, height = 6, dpi = 600)

```



frond length vs species graph 

```{r}
pussyfrond<-ggplot()+ 
  geom_violin(data=ferndata2, aes(x=Species, y=frondlength), width = 1.25)+
  theme_few()+
   theme(axis.text.x = element_text(angle = 90, size = 10, hjust=1, vjust=0.5), 
          axis.title = element_text(size=15))+
  labs(x= "Species", y= "Frond Length (mm)")+
  geom_boxplot(data=ferndata2, aes(x=Species, y=frondlength), width=0.05, color="grey20", alpha=0.2)
  

pussyfrond

ggsave("pussyfrond.png", width = 12, height = 9, dpi = 600)
```




stats for fertile frond ratio: 

```{r}
#anova for fertile frond ratio 

fertrat<- lm(fertileratio~Substrate, data = ferndata2)

par(mfrow = c(1,2))  
hist(fertrat$residuals)   # Makes histogram of residuals  
plot(fertrat, which = 2)   # Makes Q-Q plot
par(mfrow = c(1,1))
#normality is fine 

# Checking homoscedasticity 
plot(fertrat, which = 1)  # Makes residuals VS fitted plot
#kinda violated - run welch f test instead? 

summary(fertrat)
anova(fertrat)

test4<-lmer(fertileratio~Substrate+(1|Species), data=ferndata2, REML=TRUE) #make mixed effects test 
anova(test4) # run anova 
simulateResiduals(fittedModel = test4, plot=T) #check assumptions 
R2GLMER(test4)
confint(test4)
```

 
FERTILE RATIO STUFF: 

```{r}
#rock mean
exp(0.38955)
#ci
exp(0.25490189)
exp(0.6405760)


#soil mean 
exp(0.38955+0.16454)
#soilci
exp(0.25490189-0.07769897)
exp(0.6405760+0.2490294)


#wall mean 
exp(0.38955+0.27331)
#wall ci 
exp(0.25490189-0.08528648)
exp(0.6405760+0.3540057)

```

 
rock mean: 1.476316
rock ci : 1.290335, 1.897574

soil mean: 1.740357
soil ci: 1.193873, 2.434169


wall mean: 1.940334
wall ci: 1.184849,2.703593 


soil is not significantly higher than rock (T=1.83, P=0.06986), however wall is significantly higher than rock (t=2.96  p=0.00375)
 
 
 
 
figure for fertile ratio

```{r}
#new table
sum_table2 <- ferndata2 %>% group_by(Substrate) %>% summarise (mean_fertrat = mean(fertileratio), sdev_fertrat = sd(fertileratio), se_fertrat = sd(fertileratio)/sqrt(n()))

#figure 
fertilegraph<- ggplot()+
  geom_violin(data =ferndata2, aes(x=Substrate , y=fertileratio))+
  geom_boxplot(data = ferndata, aes(x=Substrate , y=fertileratio),width=0.03, color="#55B03C", alpha=0.2)+
  labs(x= "Substrate", y= "Ratio of fertile fronds to total fronds")+
  ylim(0,1)+
  theme_few()+
  theme(legend.position = "none")

fertilegraph

ggsave("fertilegraph.png", width = 9, height = 6, dpi = 600)
```



making a graph shwoing species distribution in each substrate 

```{r}
abundsp<- ferndata %>% group_by(Substrate) %>% count(Species)

abundsp<-abundsp %>% mutate(abundance=n, totalab=sum(abundance), relabund=abundance/totalab)



#CHANGE NAMES OF GROUP LABELS 
RAbar<-ggplot(data=abundsp, aes(y=relabund, x=Species, fill=Substrate))+
  geom_bar(stat="identity")+
  theme_few()+
  facet_wrap(~Substrate, ncol=1)+
  theme(axis.text.x = element_text(angle = 90, size = 10, hjust=1, vjust=0.5), 
          axis.title = element_text(size=15),
          legend.position = "none")+
  labs(y ="Relative Abundance")+
  scale_fill_manual(values=c("Wall"="#853838", "Soil"="#1A702E", "Rock"="grey30"))

RAbar

ggsave("RAbar.png", width = 9, height = 9, dpi = 600)
```




now plot for blade width: 
- im thinking a lot of this would be controlled for by species

```{r}
widthmodel<-lmer(bladewidth~Substrate + (1|Species), data=ferndata2)

anova(widthmodel)

simulateResiduals(fittedModel = widthmodel, plot=T)

# try logging data - voilates lots of assumptions 


logwidthmodel<-lmer(log_bladewidth ~Substrate +(1|Species), data=ferndata2)

summary(logwidthmodel)

R2GLMER(logwidthmodel)

simulateResiduals(fittedModel = logwidthmodel, plot=T)

confint(logwidthmodel)

```

blade width stats: 


soil is significantly higher than rock (t =2.292, p = 0.0239). wall is significantly lower than rock (t = -3.006, p = 0.0033)


```{r}
#mean rock 
exp(4.0484)
#ci
exp(3.4589705)
exp(4.6587326)

#mean soil 
exp(4.0484+ 0.4968)
#ci
exp(3.4589705+0.0765225)
exp(4.6587326+0.9297629)

#mean wall
exp(4.0484-0.8546)
#ci 
exp(3.4589705-1.4896301)
exp(4.6587326-0.2857801)


```


soil is significantly higher than rock (t =2.292, p = 0.0239). wall is significantly lower than rock (t = -3.006, p = 0.0033)

mean rock: 57.30569
ci: 31.78424, 105.5023

mean soil: 94.17926
ci: 34.31193, 267.3331

mean wall: 24.3809
ci: 7.165948, 79.27735


Tukeys test
```{r}
tukeytest3<- emmeans(logwidthmodel, list(pairwise~Substrate), adjust ="tukey")
tukeytest3
```



blade width graph: 


```{r}
#new table
widthtable <- ferndata2 %>% group_by(Substrate) %>% summarise (mean_bladewidth = mean(bladewidth), sdev_bladewidth= sd(bladewidth), se_bladewidth = sd(bladewidth)/sqrt(n()))

#figure 
frondsub<- ggplot()+
 geom_violin(data =ferndata2, aes(x=Substrate , y=bladewidth), width =1.5)+
  geom_boxplot(data = ferndata, aes(x=Substrate , y=bladewidth),width=0.03, color="#55B03C", alpha=0.2)+
  labs(x= "Substrate", y= "Blade Width (mm)")+
  theme_few()+
  theme(legend.position = "none")

frondsub
ggsave("bladewidth.png", width = 9, height = 6, dpi = 600)
```




stipe length vs substrate 


```{r}
stipemod<-lmer(log_stipelength ~ Substrate + (1|Species), data=ferndata2)
anova(stipemod)

summary(stipemod)

R2GLMER(stipemod)

simulateResiduals(fittedModel = stipemod, plot=T)

confint(stipemod)

```


stipelength stats: 

```{r}
#mean rock 
exp(3.9838)
#ci rock 
exp(3.32059476)
exp(4.6565479)


#mean soil 
exp(3.9838+0.5965)
#ci soil 
exp(3.32059476+0.08032801)
exp(4.6565479+1.1268291)


#mean wall 
exp(3.9838-1.2181)
#ci wall 
exp(3.32059476-1.92642512)
exp(4.6565479-0.5423409) 
```

mean rock: 53.72079
conf int: 27.67681, 105.272


mean soil: 97.54365
conf int: 29.99176, 324.8544


mean wall: 15.89016
conf int: 4.031625, 61.20366



figure for stipe: 

```{r}
stipetable <- ferndata2 %>% group_by(Substrate) %>% summarise (mean_stipe = mean(stipelength), sdev_stipe= sd(stipelength), se_stipe = sd(stipelength)/sqrt(n()))

#figure 
frondsub<- ggplot()+
  geom_point(data =ferndata2, aes(x=Substrate , y=stipelength), shape=21)+
  geom_point(data = ferndata2, aes(x=Substrate , y=stipelength), stat="summary", fun="mean", size=4)+
  geom_errorbar(data=stipetable, aes(x=Substrate, ymin=(mean_stipe-se_stipe), ymax=(mean_stipe+se_stipe)), width=0.1)+
  labs(x= "Substrate", y= "Stipe Length (mm)")+
  theme_few()+
  theme(legend.position = "none")

frondsub
ggsave("stipelength.png", width = 9, height = 6, dpi = 600)
```

